---
title: "Gov 50 Final Project"
author: Cielo Long 
description: "Understanding the Relationship Between Democratic Governance & Repression" 
output:
  distill::distill_article:
    self_contained: false
---

# Introduction 

Does the level of democracy in a country influence its propensity for violent repression against its citizens? In this study I explore the relationship between the level of democracy in a specific country and its engagement in political repression. Due to the multifaceted nature of repression, this study will measure repression by examining both the frequency of purges that take place within a country as well as the prevalence of political violence. I hypothesize that less democratic countries are more prone to political violence, as they are less likely to control radical political expression allowing violent extremists more room to operate, potentially leading to higher levels of civil unrest, protests, and violent conflicts. In addition, given that there is typically more opposition towards an illegitimate regime, less democratic countries will be more inclined to purge their citizens with the hope of mitigating any threats. When countries are democratic and free, they are safer and more prosperous as citizens are less likely to retaliate against the government since they have control over their lives and work. 

By understanding the dynamics between democracy and repression, I hope to explore how we as a society can increase the promotion of civil liberties and bolster the protections of citizens all across the globe. Efforts have been taken globally to promote democratic institutions of government, indicating that democratic governance reduces the likelihood of political repression. This has led to many policy makers in foreign policy to mitigate conditions that lead to repressive actions. If my hypothesis is supported, this would potentially suggest that policies promoting democratic governance may lead to the reduction of political instability and may serve as a strategy for reducing repressive actions taken against citizens. With that said, the information collected from this study could potentially be useful for civil society organizations as well as political officials who work on promoting human rights. 

# Data Section 

The data used for this cross-sectional study was derived from the Political Violence Data set. My sample is composed of a selection of countries from various regions and with varying degrees of democracy. The explanatory variable of interest is the level of democracy and the outcome variables of interest include the number of purges in a country and the number of deaths from political violence. As mentioned previously, repression is a complex phenomenon. Therefore, it is imperative that I use these two variables to provide a comprehensive analysis of political repression. While purges reflect the repressive actions taken against certain groups of the population, deaths from political violence reflects repressive actions taken against the general population as a whole. By providing different dimensions of repression, this study will offer a more nuanced understanding of how countries with varying levels of democracy engage in varying forms of repression. 

The scale of democracy is measured on a scale of 1-7 with 7 being the lowest level of democracy and 1 being the highest. This methodology was first adopted by Raymond Gastil, a Harvard-trained specialist in the field of civil liberties. Rather than using surveys or statistical models, this scale is reproduced each year using a combination of expert coding and analysis. To determine the score of democracy given to each country, in-house analysts assess a country's electoral process, political participation, functioning of government, freedom of expression etc. *Purges are measured based on* *Deaths from political violence are measured based on* If there is a significant and positive correlation between lower ranges of democracy and higher rates of purges and deaths from political violence then it would support my hypothesis as it would suggest that political violence is more common in countries that are less democratic. On the other hand, if there is no significant relationship between lower levels of democracy and higher rates of purges and deaths from political violence, this would provide evidence against my hypothesis. 

Below illustrates the two outcome variables of interest: The frequency of purges and deaths resulting from political violence 
 
# Reshaping Dataset

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
violence_data <- read.csv("~/Downloads/violence.csv", header=TRUE)
violence_data_rshp <- violence_data |>
  janitor::clean_names() |>
  rename(corrupti80 = corrupti) |>
  select(code, sftgname, starts_with(c("purges", "democ", "deaths_pc", "corrupti", "gyp", "govtcris"))) |>
  pivot_longer(cols = starts_with(c("purges", "democ", "deaths_pc", "corrupti", "gyp", "govtcris")), 
               names_to = c(".value", "decade"),
               names_pattern = "([a-z]+)([0-9]+)") |>
  rename(deaths = pc) |>
  mutate(deaths_thousand = deaths *1000) |>
  filter(!is.na(decade)) |>
  mutate(categorized_democracy = case_when(
           democ >= 6 ~ "low",
           democ < 3 ~ "high", 
           democ >= 3 & democ <= 6 ~ "middle"),
  categorized_democracy = factor(categorized_democracy, 
                                 levels = c("high", "middle", "low")),
  binary_democracy = case_when(
           democ < 5 ~ "democracy", 
           democ >= 5 ~ "autocracy"))

```

# Outcome variable (Purges) 
```{r}
purges_plot <- violence_data_rshp |>
  filter(!is.na(democ)) |>
  ggplot(aes(x = factor(decade), y= purges, fill = categorized_democracy)) +
  geom_col(position = "dodge") +
  labs(title = "Distribution of Purges by Categorized Democracy Levels",
       x = "Decade",
       y = "Number of Purges",
       legend) +
  scale_fill_manual(name = "Democracy level", values = c("low" = "blue", "middle" = "green", "high" = "red")) +
  theme_minimal()

purges_plot
```

For this visualization, I decided to categorize countries according to their level of democracy and asses how often these countries purge their citizens. The scale of democracy is measured from 1-7 with 7 being the lowest level of democracy and 1 being the highest. Using this scale, I segmented countries into "low" "middle" and "high" based on their levels of democracy. On the x axis we observe the time frame of these occurrences (70s & 80s), while the y axis represents the frequency of purges within these countries. As seen in the bar plot above, we notice that typically, less democratic countries tend to engage more frequently in the purging of their citizens. It is also noteworthy to mention that high democratic countries show no evidence of purging their citizens within during this time frame. *talk about skewed?*


# Outcome Variable: Deaths from political violence 
```{r}
purges_plot <- violence_data_rshp |>
  filter(!is.na(democ)) |>
  ggplot(aes(x = factor(decade), y= (deaths_thousand), fill = categorized_democracy)) +
  geom_col(position = "dodge") +
  labs(title = "Number of Deaths From Political Violence Per Thousand",
       x = "Decade",
       y = "Number of Deaths",
       legend) +
  scale_fill_manual(name = "Democracy level", values = c("low" = "blue", "middle" = "green", "high" = "red")) +
  theme_minimal()

purges_plot
```
This visualization represents the second key outcome variable: deaths resulting from political violence. intriguingly, we notice that in the 70s high democratic countries actually experienced more citizen deaths from political violence than those with moderate or low levels of democracy. However, it is important to note that an outlier could have very well skewed the results since we were gathering the mean of deaths. In the 80s this pattern shifts. We see that on average, countries with lower levels of democracy experience more citizen deaths from political violence in comparison to moderate or high levels of democracy. 

# Results 
 
```{r}
boxplot1<- violence_data_rshp|>
  filter(deaths_thousand > 0) |>
  filter(!is.na(democ)) |>
  ggplot(mapping = aes(x = categorized_democracy, y = log(deaths_thousand))) +
  labs(
    x = "Levels of Democracy",
    y = "Deaths from Political Violence (in M)",
    title = "The relationship between Democracy & Deaths") + 
geom_boxplot() 
boxplot1
```
Prior to performing a regression of the countries with varying levels of democracy and deaths resulting from political violence, a box plot can visualize the uni-variate distribution between these two variables. The box plot graph above plots the levels of democracy on the x-axis segmented into the categories "low" "middle" and "high", and plots the number of deaths per thousand citizens (log) on the y-axis. 

From this visualization we observe that High levels of democracy are generally associated with fewer deaths from political violence, with some notable exceptions (outliers). Middle levels of democracy have a wider spread of deaths, indicating more variability in political violence outcomes. Low levels of democracy also have a wide spread but tend to have a higher median number of deaths from political violence, which can imply that lower democracy levels are more often associated with higher deaths, although this is not universally the case.


```{r}
scatterplot1 <- ggplot(violence_data_rshp, aes(x = democ, y = purges)) +
  geom_jitter(alpha = 0.5, size = 3) +  
  geom_smooth(method = "lm", se = TRUE, color = "indianred1", size = 1) +  
  scale_color_brewer(palette = "Set1") + 
  labs(
    x = "Levels of Democracy",
    y = "Purges",
    title = "The relationship between Democracy & Purges"
  ) +
  theme_minimal() +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5), 
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14)  
  )

scatterplot1
```

Prior to performing a regression of the countries with varying levels of democracy and their frequency of purges, a scatterplot can visualize the bivariate relationship between these two variables. The box plot graph above plots the levels of democracy on the x-axis, and plots the number of purges on the y-axis. 

From this visualization we observe that as the level of democracy decreases there is an observable increase in the frequency of purges. According to the linear model used, this relationship/trend is not statistically significant due to the flat trend line. This indicates that although less democratic countries tend to purge their citizens more, there could also be other factors that determine whether these countries purge their citizens 

# Regression (deaths from poltiical violence)
```{r}
df_reg <- violence_data_rshp 

deaths_PV <- lm(deaths ~ as.factor(categorized_democracy) + corrupti + gyp  + as.factor(govtcris), data = df_reg)

summary(deaths_PV)

```
```{r}
pv_labels <- c(
    "(Intercept)" = "Intercept",
    "democ" = "Level of Democracy",
    "corrupti" = "Level of Corruption",
    "gyp" = "GDP Per Capita",
    "govtcris" = "Gov. Crisis"
  )
  modelsummary::modelsummary(deaths_PV,
                             statistic = c("s.e. = {std.error}",
                                           "p = {p.value}"),
                             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
# Interpetation of Regression

This regression indicates the difference in the number of deaths from political violence for countries with a low level of democracy compared to the high and middle levels of democracy. The coefficient is 0.000 with a standard error of 0.077 and a p-value of 1.000. This suggests that there is no difference in the number of deaths from political violence between low and high democracy countries within this model, and the result is not statistically significant.

# Statistical Significance

The association between levels of democracy and deaths resulting from political violence is not statistically significant at the alpha level of 0.05. This means that, based on our sample data and the regression model used, we do not have sufficient evidence to conclude that varying levels of democracy have a distinct effect on the number of deaths resulting from political violence. The p-values for the middle and low levels of democracy, when compared to the high level (assuming high is the reference category), are above the threshold of 0.05, indicating that any observed differences in death rates across these categories of democracy could very well be due to random chance rather than a true underlying relationship.




#Regression for Purges  
```{r}
df_reg <- violence_data_rshp 

purges_r <- lm(purges ~ as.factor(categorized_democracy) + corrupti + gyp  + as.factor(govtcris), data = df_reg)

summary(purges_r)
```
#Table for regression (purges)
```{r}
purges_labels <- c(
    "(Intercept)" = "Intercept",
    "democ" = "Level of Democracy",
    "corrupti" = "Level of Corruption",
    "gyp" = "GDP Per Capita",
    "govtcris" = "Gov. Crisis"
  )
  modelsummary::modelsummary(purges_r,
                             statistic = c("s.e. = {std.error}",
                                           "p = {p.value}"),
                      
                             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
# Interpretation for regression 
This regression indicates that countries lower in democratic levels will  while holding other variables constant. Note that the scale of democracy is measured from 1-7 with 7 being the lowest level of democracy and 1 being the highest. With that said, a one unit "decrease" in democracy will result in a increase in the frequency of purges.  

 
# Statistical Significane 

While holding other variables constant, The coefficient for the level of democracy is 0.037, suggesting a very small positive association with the frequency of purges; however, this is not statistically significant (p = 0.270), with a standard error of 0.033.

# Causality and confounders 

Since the explanatory variable of the 

# Conclusion Section 
The regression results in this study contradict one portion of my analysis which is that less democratic countries experience more deaths as a result from political violence. Nonetheless, I still believe that it is important to note that the porti

Notes: 
- don't want code to appear
- what if both of my results are not statistically significant? do i find the confidence intervals?